ECCos from the Black Box: Letting Models speak for Themselves

### 2023-04-24 paper/paper.tex
paper updates

\title{ECCoEs from the Black Box: Letting Models speak for Themselves}

@@

\begin{abstract}
  We propose Eccoe: an effortless and rigorous way to produce plausible and conformal Counterfactual Explanations for Black Box Models using Conformal Prediction. To address the need for plausible explanations, existing work has primarily relied on surrogate models to learn the data-generating process. This effectively reallocates the task of learning realistic representations of the data from the model itself to the surrogate. Consequently, the generated explanations may look plausible to humans but not necessarily conform with the behaviour of the Black Box Model. We formalise this notion through the introduction of new evaluation measures. In order to still address the need for plausibility, we build on a recent approach that works by minimizing predictive model uncertainty. Using differentiable Conformal Prediction, we relax the previous assumption that the Black Box Model can produce predictive uncertainty estimates.
\end{abstract}

@@

Counterfactual Explanations are a powerful, flexible and intuitive way to not only explain Black Box Models but also enable affected individuals to challenge them through the means of Algorithmic Recourse. Instead of opening the black box, Counterfactual Explanations work under the premise of strategically perturbing model inputs to understand model behaviour \citep{wachter2017counterfactual}. Intuitively speaking, we generate explanations in this context by asking simple what-if questions of the following nature: `Our credit risk model currently predicts that this individual's credit profile is too risky to offer them a loan. What if they reduced their monthly expenditures by 10\%? Will our model then predict that the individual is credit-worthy'?

@@

\section{From Adversarial Examples to Plausible Explanations}\label{background}

@@

@@

@@

@@

\section{Evaluating the Faithfulness of Counterfactuals}\label{conformity}

@@

Fortunately, recent work by \citet{grathwohl2020your} on Energy Based Models (EBM) has pointed out that there is a `generative model hidden within every standard discriminative model'. The authors show that we can draw samples from the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y})$ using Stochastic Gradient Langevin Dynamics (SGLD). In doing so, it is possible to train classifiers jointly for the discriminative task using standard cross-entropy and the generative task using SGLD. They demonstrate empirically that among other things this improves predictive uncertainty quantification for discriminative models.

@@

\section{A Framework for Conformal Counterfactual Explanations}\label{cce}

Now that we have a framework for evaluating Counterfactual Explanations in terms of their plausibility and conformity, we are interested in finding a way to generate counterfactuals that are as plausible and conformal as possible. We hypothesize that a narrow focus on plausibility may come at the cost of reduced conformity. Using a surrogate model for the generative task, for example, may improve plausibility but inadvertently yield counterfactuals that are more consistent with the surrogate than the Black Box Model itself. We suggest that one way to ensure model conformity is to rely strictly on the model itself. In this section, we introduce a novel framework that meets this requirement, works under minimal assumptions and does not impede the plausibility objective: Conformal Counterfactual Explanations.

\subsection{Plausible Counterfactuals through Minimal Uncertainty}

Our proposed methodology is built on the findings presented in~\citet{schut2021generating}. The authors demonstrate that it is not only possible but remarkably easy to generate plausible counterfactuals for Black Box Models that provide predictive uncertainty estimates. Their proposed algorithm solves Equation~\ref{eq:general} by greedily applying JSMA in the feature space with standard cross-entropy loss and no penalty at all. They show that this is equivalent to minimizing predictive uncertainty and hence yields counterfactuals for which the model $M_{\theta}$ predicts the target label $\mathbf{y}^*$ with high confidence. Provided the model is well-calibrated, these counterfactuals are plausible which the authors demonstrate empirically through benchmarks \citep{schut2021generating}.

Unfortunately, this idea hinges on the crucial assumption that the Black Box Model provides predictive uncertainty estimates. The authors argue that in light of rapid advances in Bayesian Deep Learning (DL), this assumption is overall less costly than the engineering overhead induced by using surrogate models. This is even more true today, as recent work has put Laplace Approximation back on the map for truly effortless Bayesian DL \citep{immer2020improving,daxberger2021laplace,antoran2023sampling}. Nonetheless, the need for Bayesian methods may be too restrictive in some cases.

In looking for ways to lift that restriction, we found a promising alternative candidate for predictive uncertainty quantification (UQ) that we will briefly introduce next: Conformal Prediction.

\subsection{Conformal Prediction}

@@

>>>>

\title{ECCos from the Black Box: Letting Models speak for Themselves}

@@

\begin{abstract}
  Counterfactual Explanations offer an intuitive and straightforward way to explain Black Box Models but they are not unique. To address the need for plausible explanations, existing work has primarily relied on surrogate models to learn how the input data is distributed. This effectively reallocates the task of learning realistic representations of the data from the model itself to the surrogate. Consequently, the generated explanations may look plausible to humans but not necessarily conform with the behaviour of the Black Box Model. We formalise this notion of model conformity through the introduction of tailored evaluation measures and propose a novel algorithmic framework for generating \textbf{E}nergy-Constrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits. To do so, \textbf{ECCo} leverages recent advances in energy-based modelling and predictive uncertainty quantification through conformal inference. Through illustrative examples and extensive empirical studies, we demonstrate that ECCos reconcile the need for plausibility and model conformity. 
\end{abstract}

@@

Counterfactual Explanations provide a powerful, flexible and intuitive way to not only explain Black Box Models but also enable affected individuals to challenge them through the means of Algorithmic Recourse. Instead of opening the black box, Counterfactual Explanations work under the premise of strategically perturbing model inputs to understand model behaviour \citep{wachter2017counterfactual}. Intuitively speaking, we generate explanations in this context by asking simple what-if questions of the following nature: `Our credit risk model currently predicts that this individual's credit profile is too risky to offer them a loan. What if they reduced their monthly expenditures by 10\%? Will our model then predict that the individual is credit-worthy'?

@@

\section{Background and Related Work}\label{background}

@@

\subsection{Gradient-Based Counterfactual Search}\label{gradient}

@@

\subsection{From Adversial Examples to Plausible Explanations}

@@

\subsubsection{Plausibility through Surrogates}

@@

\subsubsection{Plausibility through Minimal Predictive Uncertainty}

\citet{schut2021generating} show that to meet the plausibility objective we need not explicitly model the input distribution. Pointing to the undesirable engineering overhead induced by surrogate models, they propose that we rely on the implicit minimisation of predictive uncertainty instead. Their proposed methodology solves Equation~\ref{eq:general} by greedily applying JSMA in the feature space with standard cross-entropy loss and no penalty at all. They demonstrate theoretically and empirically that their approach yields counterfactuals for which the model $M_{\theta}$ predicts the target label $\mathbf{y}^*$ with high confidence. Provided the model is well-specified, these counterfactuals are plausible.

Unfortunately, this idea hinges on the assumption that the Black Box Model provides well-calibrated predictive uncertainty estimates. The authors argue that in light of rapid advances in Bayesian Deep Learning (DL), this assumption is overall less costly than the engineering overhead induced by using surrogate models. This is even more true today, as recent work has put Laplace Approximation back on the map for truly effortless Bayesian DL \citep{immer2020improving,daxberger2021laplace,antoran2023sampling}. Nonetheless, the need for Bayesian methods may be too restrictive in some cases. Our proposed methodology, which we will turn to next, relaxes this restriction.

\section{Energy-Constrained Conformal Counterfactuals}\label{ecco}

The primary objective of this work has been to develop a methodology for generating maximally plausible counterfactuals under minimal intervention. Our proposed framework is based on the premise that explanations should be plausible but not plausible at all costs. Energy-Constrained Conformal Counterfactuals (ECCo) achieve this goal in two ways: firstly, they rely on the Black Box itself for the generative task; and, secondly, they involve an approach to predictive uncertainty quantification that is model-agnostic.

\subsection{Quantifying the Model's Generative Property}

@@

Recent work by \citet{grathwohl2020your} on Energy Based Models (EBM) has pointed out that there is a `generative model hidden within every standard discriminative model'. The authors show that we can draw samples from the posterior conditional distribution $p_{\theta}(\mathbf{x}|\mathbf{y})$ using Stochastic Gradient Langevin Dynamics (SGLD). In doing so, it is possible to train classifiers jointly for the discriminative task using standard cross-entropy and the generative task using SGLD. They demonstrate empirically that among other things this improves predictive uncertainty quantification for discriminative models.

@@

\subsection{Quantifying the Model's Predictive Uncertainty}

@@

\section{Evaluation Framework}\label{conformity}
